Pipit: Reactive Systems in \fstar ~(Extended Abstract)

### 2023-05-16 2023pipit/Main.tex
copyediting

\begin{abstract}
  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
  Pipit can also generate imperative code in \lowstar{}, a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
\end{abstract}

@@

Safety-critical control systems, such as the anti-lock braking systems that occur in most cars today, need to be correct and execute in real-time.
One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
This mismatch may occur even when the compiler has been verified correct, as in the case of Vélus~\cite{bourke2017formally}.
For example, in the Vélus compiler, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB}.

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on a clock-free subset of Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers, and using \lowstar{}'s C-code generation for real-time execution.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
Pipit translates its expression language to a transition system for k-inductive proofs, which is verified\footnote{The translation is verified for a previous untyped expression representation; we are still finishing porting the proofs to the new typed representation.}; verifying the translation to imperative code is ongoing.

@@

% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
The coefficient vector is represented by a list of reals here, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
The implementation starts by looking at the list of coefficients, and returns zero if the list is empty.
If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before finally summing the two parts.
At the start of execution the delay is initialised to zero.

Although this is a simple definition, verifying it \emph{and} executing it in Lustre is somewhat impractical, as tool support varies widely.
To write this program in Lustre, we would likely need to reformulate the program to use arrays.
Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays without compiler support~\cite{champion2016kind2}; and Lustre V6 does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.

Instead, inspired by Bedrock~\cite{chlipala2013bedrock}, we define Pipit with a small core language embedded inside a theorem prover (\fstar{} in our case).
By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program without lists or arrays.
% which can be easily executed in statically-bounded runtime and space.
For example, we can compute the weighted moving average with the following concrete coefficients:

@@

Normalising the above definition of $\mbox{fir}_2$ removes all lists.
The result fits in the core language defined in \autoref{s:core}, for which we can generate statically-allocated imperative code:

@@

This property states that
% for the coefficients used by $\fir_2$,
  if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
To prove that this property holds, we translate to a transition system and show that the stream is always true.
In this case, induction over the transition relation is sufficient to prove the property.
There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:

@@

If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is no longer sufficient; we need a stronger invariant.
One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:

@@

The grammar of Pipit is defined in \autoref{f:core-grammar}.
The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions other than reusing closed functions from the \fstar{} meta-language.
Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
Streams can be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the latest values from stream $e'$ for subsequent steps.

@@

% Here, \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
As with the examples in \autoref{ss:tut:bibo}, the above properties can be automatically verified.
% by \fstar{}'s proof automation.
Pipit generates real-time C code for this example.

>>>>

\begin{abstract}
  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
  Pipit can also generate imperative code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
\end{abstract}

@@

Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
This mismatch may occur even when the compiler has been verified correct, as in the case of Vélus~\cite{bourke2017formally}.
For example, in the Vélus compiler, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB}.
% TODO: link to Kind2 issue?

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on a subset of Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.

@@

% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
The implementation starts by looking at the list of coefficients, and returns zero if the list is empty.
If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before finally summing the two parts.
At the start of execution the delay is initialised to zero.

Although this is a simple definition, verifying it \emph{and} executing it in Lustre is somewhat impractical, as tool support varies widely.
To write this program in Lustre, we would likely need to reformulate the program to use arrays.
Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays without compiler support~\cite{champion2016kind2}; and Lustre V6 does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
%TODO defer explanation to related work

Instead, inspired by Bedrock~\cite{chlipala2013bedrock}, we define Pipit as a small core language embedded inside a theorem prover (\fstar{} in our case).
By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program without lists or arrays.
For example, we can compute the weighted moving average with the following concrete coefficients:

@@

Normalising the above definition of $\mbox{fir}_2$ removes all lists.
The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:

@@

This property states that if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
To prove that this property holds, we translate to a transition system and show that the stream is always true.
In this case, induction over the transition relation is sufficient to prove the property.
There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:

@@

If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is no longer sufficient; we need a stronger invariant.
One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
% TODO: intuition, (0 fby x) < n = true fby (x < n)

@@

The grammar of Pipit is defined in \autoref{f:core-grammar}.
The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(\mlamX{})$.
Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
Streams can be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the latest values from stream $e'$ for subsequent steps.

@@

% Here, \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
As with the examples in \autoref{ss:tut:bibo}, the two properties can be automatically verified.
% by \fstar{}'s proof automation.
Pipit generates real-time C code for this example.
